---
title: "Introduction"
author: "George Wiggins"
date: "11/12/2020"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
    theme: lumen
    css: www/css/master.css
    highlight: default
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA, fig.align = 'center')
```

# Datasets

To illustrate some commonly statistical procedures in biology a few near-real world dataset will be used.  

## Knockdown of candidate gene
This dataset is based on the hypothesis that the deletion of a gene (geneX) will alter the growth of endometrial cancer cells. To test this, two cell lines (AN3CA and EM-E6/E7-hTERT) were transfected with one of three siRNA, all targeting geneX. After transfection, proliferation (cell metabolism) of each experimental condition was measured by a MTT assay. These experiments were repeated in triplicate across three weeks.

## Chick weights on different diets
This is a dataset that is avaliable through an R package `chickwts` that measure the weight of different chicks on different feeds. 

## Novel therapies: blood haemoglobin
This is a hypothetical dataset made to compare several treatments to standards care. The change in blood haemoglobin from baseline was recorded for 10 individuals randomly assigned to either standard care or one of six novel therapies (named DrugA-DrugF).

# R 
This is a desgined to be a simple guide to performing statistical analysis in R. There will be some statistical theory but the purpose is to show how statistical tests can be executed in R and how to display these results. 

For those not familiar with R, R is a statistical language which is useful for more complex statistical procedures and presenting data. For the most part, this guide will have provide simple examples and "how-tos". A major challenge is often "wrangling" data into a format so that a function can be executed. For data wrangling, I will try use the latest `tidyverse` syntax. Alternatively, it is possible to 'wrangle'/organise data in excel before importing into R.

# To start
I recommend installing [RStudio Desktop](https://www.rstudio.com/products/rstudio/download/), a IDE (integrated development environment) for R that allows a more user friendly experience of R. If you prefer writing in a console, then you can install [R](https://www.r-project.org/). 

Here are some packages that are used to generate this file, some may need to be installed through CRAN using the `install.packages()` function, github using the `install_github()` function from `devtools` or bioconductor using `install()` from the `BiocManager` package.

```{r datasets, results='hide',error=FALSE, warning=FALSE, message=FALSE, eval=F}
## Read excelsheets
library(openxlsx)

## Packages used for 'data wrangling'
library(tidyverse)
library(data.table)

## Extra stats package
library(rstatix) ## used to help format stats into dfs for ggplot
library(DescTools) ## For Dunnett's test

## Packages for plotting and table outputs
library(ggplot2)
library(ggsci)
library(ggpubr)
library(cowplot)
library(kableExtra) ## used for to make nicer looking tables
```

## Tidyverse syntax
Tidyverse (`tidyr`, `dpylr`) has a unique syntax/funcationality compared to base R that allows the output of one function to be 'piped' into a second function. Piping is implemented with by the pipe function (`%>%`). As a simple explaination, lets assume a user whats to exclude rows based on some term and then summarise that data. This can be achieved by piping, or by storing the ouputs of each function, I'll use the mtcars dataset to demonstrate. 
```{r tidyr}
library(tidyverse)
## Tidyverse with piping
group_by(mtcars, cyl) %>% filter(cyl != 4) %>% summarise(mean=mean(mpg))
```

```{r baseR}
## No piping
x <- group_by(mtcars, cyl)
y <- filter(x, cyl !=4)
summarise(y, mean=mean(mpg))
```

Note: The function I used are all from the tidyverse world, this just illustrates the piping function that will be use regularly throughout the document.

**Importing data**

R can phrase several file types, the easiest are flat files (e.g. tab-delimited text files or comma-separated values files). I have uploaded several files to a github which you can download and import them into R. As most experimental data collected is often stored in excel, I have used a package `openxlsx` which has the ability to read individual excel sheets. 


```{r ReadData, eval=F}
## Ensure the path to data is correct
sheetNames<-getSheetNames("Data/DummyData_MTT.xlsx") ## Gives you a list of sheetNames in a given excel workbook

## Read a single sheet (the first sheet) from a workbook 
# for details on function type ?function (e.g. ?read.xlsx)
Exp1 <- read.xlsx("Data/DummyData_MTT.xlsx", sheet=sheetNames[1]) # [1] denotes which sheet to read in

## Read all the sheets and store as a list
KD_data<-lapply(sheetNames,function(i) read.xlsx("Data/DummyData_MTT.xlsx", sheet=i))
names(KD_data) <- c(paste("Exp",1:3, sep=""), 'normalised')
```