---
title: "Regression"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
    theme: lumen
    css: www/css/master.css
    highlight: default
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA, fig.align = 'center')
```


```{r datasets, results='hide',error=FALSE, warning=FALSE, message=FALSE,echo=F}
## Read excelsheets
library(openxlsx)

## Packages used for 'data wrangling'
library(tidyverse)
library(data.table)

## Extra stats package
library(rstatix) ## used to help format stats into dfs for ggplot
library(DescTools) ## For Dunnett's test
library(epitools)
library(epiR)
library(mlbench)
library(pROC) ## For ROC curves

## Packages for plotting and table outputs
library(ggplot2)
library(ggsci)
library(ggpubr)
library(ggrepel)
library(cowplot)
library(kableExtra) ## used for to make nicer looking tables
```


```{r global_functions, echo=FALSE}
bxplot <- function(...){ggplot(...)+geom_boxplot(width=0.4, size=1, fatten=1, colour="grey70") +
  geom_point(colour="steelblue", size=1.2, alpha=0.5) +
  theme_bw(base_size=14)}

p.siginf<- function(vec){
  out<-c()
  for(i in 1:length(vec)){
    if(vec[i] < 0.001){
      out <- c(out, "***")
    }else if( vec[i] < 0.01){
      out <- c(out, "**")
    } else if(vec[i] <0.05){
      out <- c(out, "*")
    }else{
      out <- c(out, "ns")
    }
  }     
  return(out)
}
```



# Logistic Regression

Logistic regression is a method used to model binary outcomes. In R logistic regression is a generalised linear model and can be implemented with the `glm()` function. The `glm()` funciton has many types of models, to use logistic regression the `family` options should be set to `'binomial'` (e.g. `glm(response variable ~ predictor variables,family='binomial', data=df)`). 

```{r}
data("PimaIndiansDiabetes2")
Diabetes<- na.omit(PimaIndiansDiabetes2) # remove missing data
#levels(Diabetes$diabetes) <- 0:1 # relabel neg and pos to 0 and 1

## Fitting logistic regression predicted by all variables 
model1 <- glm (diabetes~., data=Diabetes, family = 'binomial')
## not the period tells R to use all columns other than diabetes. This is the same as
##  glm(diabetes ~ pregnant + glucose + pressure + triceps + insulin + mass + pedigree + age ,data=Diabetes, family = 'binomial')

summary(model1)
```

The summary depicts the estimate weight each term has on the prediction of diabetes. For example, with every increase unit increase of glucose will increase the log odds of diabetes by 0.036. 

